---
title: "R Notebook"
output: html_notebook
---

```{r}
install.packages("FNN")

```


# Data Cleaning

##Play Types
```{r}
library(readr)
library(tidyverse)

# Play types DF
df <- read_csv("~/Documents/Dev Projects/Sports/nba-archetypes-project/data/NBA_Play_Types_16_24.csv")

df <- df %>%
  mutate(
    FREQ = as.numeric(FREQ),
    FREQ_PCTL = as.numeric(FREQ_PCTL),
    PPP = as.numeric(PPP),
    PPP_PCTL = as.numeric(PPP_PCTL)
  )
```

```{r}
df_sub <- df %>%
  select(SEASON, PLAYER_ID, PLAYER_NAME, PLAY_TYPE,
         FREQ, FREQ_PCTL, PPP, PPP_PCTL)
head(df_sub)
```

```{r}
df_sub_clean <- df_sub %>%
  group_by(SEASON, PLAYER_ID, PLAYER_NAME, PLAY_TYPE) %>%
  summarise(
    FREQ = mean(FREQ, na.rm = TRUE),
    FREQ_PCTL = mean(FREQ_PCTL, na.rm = TRUE),
    PPP = mean(PPP, na.rm = TRUE),
    PPP_PCTL = mean(PPP_PCTL, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
# long to wide format
df_wide <- df_sub_clean %>%
  pivot_wider(
    id_cols = c(SEASON, PLAYER_ID, PLAYER_NAME),          
    names_from = PLAY_TYPE,                               
    values_from = c(FREQ, FREQ_PCTL, PPP, PPP_PCTL),      
    names_glue = "{PLAY_TYPE}_{.value}"                   
  )
tail(df_wide, 100)

```

##BBall Ref
```{r}
# Read the CSV exported from Python
df_stats <- read_csv("~/Documents/Dev Projects/Sports/nba-archetypes-project/all_players_seasons.csv")
```

```{r}
df_stats <- df_stats %>%
  rename(
    PLAYER_NAME = Player,  # match df_wide column
    SEASON = Season
  )

head(df_stats)
```

```{r}
df_merged <- df_wide %>%
  left_join(df_stats, by = c("PLAYER_NAME", "SEASON"))
tail(df_merged, 100)
```

```{r}
colnames(df_merged)
```

```{r}
df_merged_clean <- df_merged %>%
  select(-c(
    "Cut_FREQ_PCTL", "Misc_FREQ_PCTL", "PnR Roll Man_FREQ_PCTL", "Postup_FREQ_PCTL",
    "Spotup_FREQ_PCTL", "Transition_FREQ_PCTL", "Handoff_FREQ_PCTL", "Isolation_FREQ_PCTL",
    "Off Screens_FREQ_PCTL", "PnR Ball-Handler_FREQ_PCTL", "Putbacks_FREQ_PCTL",
    "Cut_PPP_PCTL", "Misc_PPP_PCTL", "PnR Roll Man_PPP_PCTL", "Postup_PPP_PCTL",
    "Spotup_PPP_PCTL", "Transition_PPP_PCTL", "Handoff_PPP_PCTL", "Isolation_PPP_PCTL",
    "Off Screens_PPP_PCTL", "PnR Ball-Handler_PPP_PCTL", "Putbacks_PPP_PCTL",
    "TRB%", "G"
  ))
colnames(df_merged_clean)
```



# K-Means
```{r}
library(dplyr)

# Select only numeric columns to cluster on
df_numeric <- df_merged_clean %>%
  select(where(is.numeric)) %>%
  select(-PLAYER_ID)  # remove identifier columns if needed

# Save player-season info to merge back later
player_info <- df_merged_clean %>%
  select(SEASON, PLAYER_NAME, PLAYER_ID)

# If NA, use mean of that column (Mess with this !!)
df_numeric_clean <- df_numeric %>%
  mutate(across(everything(), ~ ifelse(is.finite(.), ., mean(., na.rm = TRUE))))

# Normalize the numeric values
df_scaled <- scale(df_numeric_clean)
```

```{r}
set.seed(42)
k <- 8
kmeans_result <- kmeans(df_scaled, centers = k, nstart = 25, iter.max = 100)
```

```{r}
df_archetypes <- bind_cols(player_info, as.data.frame(df_scaled)) %>%
  mutate(Archetype = as.factor(kmeans_result$cluster))
```

```{r}
df_archetypes %>%
  count(Archetype) %>%
  arrange(desc(n))
```
```{r}
df_archetypes %>%
  select(SEASON, PLAYER_NAME, Archetype) %>%
  arrange(Archetype, SEASON, PLAYER_NAME)

```

```{r}
df_archetypes %>%
  group_by(Archetype) %>%
  summarise(Players = paste0(unique(PLAYER_NAME), collapse = ", "), .groups = "drop")

```


```{r}
library(Rtsne)

# Run t-SNE on the scaled data (you may want to set a smaller `perplexity` if data is small)
set.seed(42)
tsne_result <- Rtsne(df_scaled, dims = 2, perplexity = 30)

# Build dataframe for plotting
tsne_df <- as.data.frame(tsne_result$Y) %>%
  setNames(c("D1", "D2")) %>%
  bind_cols(player_info) %>%
  mutate(Archetype = as.factor(kmeans_result$cluster))

# Plot
ggplot(tsne_df, aes(x = Dim1, y = Dim2, color = Archetype)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "t-SNE Projection of NBA Player Archetypes") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "right")
```

```{r}
# Run PCA on the scaled features
pca_result <- prcomp(df_scaled, center = TRUE, scale. = TRUE)

# Combine PCA output with player info and archetypes
pca_df <- as.data.frame(pca_result$x[, 1:2]) %>%  # Take first 2 PCs
  bind_cols(player_info) %>%
  mutate(Archetype = as.factor(kmeans_result$cluster))

# Plot PCA
ggplot(pca_df, aes(x = PC1, y = PC2, color = Archetype)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "PCA Projection of NBA Player Archetypes",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "right")
```

```{r}
library(dplyr)

top5_by_cluster_pts_kmeans <- df_archetypes %>%
  left_join(df_merged_clean %>%
              select(SEASON, PLAYER_NAME, PTS),
            by = c("SEASON", "PLAYER_NAME")) %>%
  filter(SEASON == "2023-24") %>%
  group_by(Archetype) %>%
  slice_max(order_by = PTS, n = 10) %>%
  ungroup() %>%
  arrange(Archetype, desc(PTS)) %>%
  select(SEASON, PLAYER_NAME, Archetype)

top5_by_cluster_pts_kmeans
```


## Finding Most Similar Players

```{r}
df_with_scaled <- bind_cols(
  df_archetypes %>% select(PLAYER_NAME, SEASON, Archetype),
  as.data.frame(df_scaled)
)
```

```{r}
library(FNN)

find_similar_players <- function(player_name, season, df_with_scaled, df_scaled, k = 5) {
  # Step 1: Locate the target row
  target_row <- df_with_scaled %>%
    filter(PLAYER_NAME == player_name, SEASON == season)
  
  if (nrow(target_row) != 1) {
    stop("Player-season not found or not unique.")
  }
  
  # Step 2: Get target's archetype
  target_archetype <- target_row$Archetype
  
  # Step 3: Subset to same archetype
  df_same_archetype <- df_with_scaled %>%
    filter(Archetype == target_archetype)
  
  # Extract feature columns from same-archetype set
  features_same_archetype <- df_same_archetype %>%
    select(-PLAYER_NAME, -SEASON, -Archetype)
  
  # Extract feature columns for the target player
  target_features <- target_row %>%
    select(-PLAYER_NAME, -SEASON, -Archetype)
  
  # Step 4: Run KNN (k+1 to include target themselves)
  knn_result <- get.knnx(
    data = features_same_archetype,
    query = target_features,
    k = k + 1
  )
  
  # Step 5: Extract neighbors, remove self if present
  neighbor_indices <- knn_result$nn.index[1,]
  neighbors <- df_same_archetype[neighbor_indices, ]
  
  # Remove the query player-season from the output
  neighbors <- neighbors %>%
    filter(!(PLAYER_NAME == player_name & SEASON == season)) %>%
    head(k)  # ensure only k others are returned
  
  return(neighbors)
}
```

```{r}
name = "Duncan Robinson"
season = "2020-21"
n = 10

find_similar_players(name, season, df_with_scaled, df_scaled, k = n)
```




# Seeded K-Means
## Model
To try to get better clusters/ archetypes, I propose a seeded k-means method where I create 9 archetypes, and run a semi-supervised k-means algorithm, with 10 players already put into each cluster.

```{r}
library(dplyr)
library(FNN)

# Step 1: Seed player-season definitions
seeds <- tribble(
  ~PLAYER_NAME, ~SEASON, ~ARCHETYPE,
  
  # Post Scorer
  "Nikola Jokić", "2022-23", "Post Scorer",
  "Nikola Jokić", "2020-21", "Post Scorer",
  "Domantas Sabonis", "2023-24", "Post Scorer",
  "Domantas Sabonis", "2021-22", "Post Scorer",
  "Joel Embiid", "2021-22", "Post Scorer",
  "Joel Embiid", "2023-24", "Post Scorer",
  "Nikola Vucevic", "2019-20", "Post Scorer",
  "DeMarcus Cousins", "2016-17", "Post Scorer",
  "Karl-Anthony Towns", "2019-20", "Post Scorer",
  "LaMarcus Aldridge", "2018-19", "Post Scorer",

  # Stretch Big
  "Karl-Anthony Towns", "2023-24", "Stretch Big",
  "Brook Lopez", "2020-21", "Stretch Big",
  "Myles Turner", "2023-24", "Stretch Big",
  "Kevin Love", "2016-17", "Stretch Big",
  "Nemanja Bjelica", "2018-19", "Stretch Big",
  "Naz Reid", "2023-24", "Stretch Big",
  "Al Horford", "2023-24", "Stretch Big",
  "Chet Holmgren", "2023-24", "Stretch Big",
  "Wendell Carter Jr.", "2022-23", "Stretch Big",
  "Kristaps Porzingis", "2023-24", "Stretch Big",

  # Roll & Cut Big
  "Rudy Gobert", "2016-17", "Roll & Cut Big",
  "Jarrett Allen", "2023-24", "Roll & Cut Big",
  "Nic Claxton", "2022-23", "Roll & Cut Big",
  "Clint Capela", "2022-23", "Roll & Cut Big",
  "Hassan Whiteside", "2019-20", "Roll & Cut Big",
  "Derrick Favors", "2018-19", "Roll & Cut Big",
  "Steven Adams", "2015-16", "Roll & Cut Big",
  "Daniel Gafford", "2023-24", "Roll & Cut Big",
  "Onyeka Okongwu", "2023-24", "Roll & Cut Big",
  "Kevon Looney", "2021-22", "Roll & Cut Big",

  # Shot Creator
  "Russell Westbrook", "2016-17", "Shot Creator",
  "Shai Gilgeous-Alexander", "2023-24", "Shot Creator",
  "Giannis Antetokounmpo", "2017-18", "Shot Creator",
  "Luka Doncic", "2023-24", "Shot Creator",
  "Kevin Durant", "2021-22", "Shot Creator",
  "Paul George", "2016-17", "Shot Creator",
  "Anthony Edwards", "2021-22", "Shot Creator",
  "Andrew Wiggins", "2017-18", "Shot Creator",
  "Brandon Ingram", "2021-22", "Shot Creator",
  "Kobe Bryant", "2015-16", "Shot Creator",

  # Primary Ball-Handler
  "Kyle Lowry", "2017-18", "Primary Ball-Handler",
  "Tyrese Haliburton", "2021-22", "Primary Ball-Handler",
  "Chris Paul", "2022-23", "Primary Ball-Handler",
  "Kemba Walker", "2017-18", "Primary Ball-Handler",
  "Eric Bledsoe", "2018-19", "Primary Ball-Handler",
  "Ricky Rubio", "2017-18", "Primary Ball-Handler",
  "Darius Garland", "2022-23", "Primary Ball-Handler",
  "George Hill", "2015-16", "Primary Ball-Handler",
  "Josh Giddey", "2023-24", "Primary Ball-Handler",
  "Malcolm Brogdon", "2017-18", "Primary Ball-Handler",

  # Slasher
  "De'Aaron Fox", "2018-19", "Slasher",
  "John Wall", "2017-18", "Slasher",
  "Collin Sexton", "2018-19", "Slasher",
  "Jalen Williams", "2022-23", "Slasher",
  "Jonathon Simmons", "2017-18", "Slasher",
  "Elfrid Payton", "2020-21", "Slasher",
  "Emmanuel Mudiay", "2018-19", "Slasher",
  "Markelle Fultz", "2023-24", "Slasher",
  "Delon Wright", "2017-18", "Slasher",
  "RJ Barrett", "2021-22", "Slasher",

  # Finisher
  "Michael Kidd-Gilchrist", "2016-17", "Finisher",
  "Aaron Gordon", "2016-17", "Finisher",
  "Herbert Jones", "2021-22", "Finisher",
  "Christian Braun", "2022-23", "Finisher",
  "Pascal Siakam", "2017-18", "Finisher",
  "Hamidou Diallo", "2022-23", "Finisher",
  "Bruce Brown", "2020-21", "Finisher",
  "Kyle Anderson", "2016-17", "Finisher",
  "Ausar Thompson", "2023-24", "Finisher",
  "Tari Eason", "2022-23", "Finisher",

  # Movement Shooter
  "JJ Redick", "2016-17", "Movement Shooter",
  "Klay Thompson", "2017-18", "Movement Shooter",
  "Kyle Korver", "2017-18", "Movement Shooter",
  "Duncan Robinson", "2020-21", "Movement Shooter",
  "Buddy Hield", "2022-23", "Movement Shooter",
  "Sam Hauser", "2023-24", "Movement Shooter",
  "Max Strus", "2022-23", "Movement Shooter",
  "Joe Harris", "2020-21", "Movement Shooter",
  "Marco Belinelli", "2017-18", "Movement Shooter",
  "Kentavious Caldwell-Pope", "2018-19", "Movement Shooter",

  # Stationary Shooter
  "Channing Frye", "2016-17", "Stationary Shooter",
  "P.J. Tucker", "2019-20", "Stationary Shooter",
  "Nicolas Batum", "2020-21", "Stationary Shooter",
  "Otto Porter Jr.", "2021-22", "Stationary Shooter",
  "Dorian Finney-Smith", "2021-22", "Stationary Shooter",
  "Trevor Ariza", "2017-18", "Stationary Shooter",
  "Danuel House Jr.", "2019-20", "Stationary Shooter",
  "Jae Crowder", "2018-19", "Stationary Shooter",
  "Grant Williams", "2022-23", "Stationary Shooter",
  "John Konchar", "2022-23", "Stationary Shooter",

  # 2-Level Playmaker
  "Alex Caruso", "2023-24", "2-Level Playmaker",
  "Tobias Harris", "2019-20", "2-Level Playmaker",
  "OG Anunoby", "2017-18", "2-Level Playmaker",
  "Bogdan Bogdanovic", "2019-20", "2-Level Playmaker",
  "Patrick Beverly", "2019-20", "2-Level Playmaker",
  "Marcus Smart", "2018-19", "2-Level Playmaker",
  "Keldon Johnson", "2022-23", "2-Level Playmaker",
  "Will Barton", "2017-18", "2-Level Playmaker",
  "Kelly Oubre", "2023-24", "2-Level Playmaker",
  "Luguentz Dort", "2020-21", "2-Level Playmaker"
)

df_with_scaled_mutated <- df_with_scaled %>%
  mutate(
    `ORB%` = `ORB%` * 1.5,
    `Postup_FREQ` = `Postup_FREQ` * 1.5,
    `3P%FGA` = `3P%FGA` * 1.5
  )

# Step 2: Join with scaled features
seed_features <- df_with_scaled_mutated %>%
  inner_join(seeds, by = c("PLAYER_NAME", "SEASON"))

# Step 3: Compute mean vector for each archetype
initial_centers <- seed_features %>%
  group_by(ARCHETYPE) %>%
  summarise(across(where(is.numeric), mean), .groups = "drop")

# Step 4: Run seeded k-means
set.seed(38)
kmeans_seeded <- kmeans(
  x = df_with_scaled_mutated %>% select(where(is.numeric)),
  centers = as.matrix(initial_centers %>% select(-ARCHETYPE)),
  iter.max = 50, nstart = 1
)

# Step 5: Add cluster assignments back
df_with_clusters <- df_with_scaled_mutated %>%
  mutate(Cluster = kmeans_seeded$cluster)

```

```{r}
df_with_clusters_named %>%
  count(ARCHETYPE, Cluster) %>%
  arrange(Cluster)
```

## Top 10 per Cluster
```{r}
library(purrr)

# Step 1: Get seed player identifiers
seed_keys <- seeds %>% select(PLAYER_NAME, SEASON)

# Step 2: Compute distance to assigned cluster center
centers <- kmeans_seeded$centers
features_only <- df_with_scaled_mutated %>% select(where(is.numeric))

distances <- map_dbl(1:nrow(features_only), function(i) {
  clust <- kmeans_seeded$cluster[i]
  sum((features_only[i, ] - centers[clust, ])^2)
})

# Step 3: Add cluster + distance
df_with_clusters <- df_with_clusters %>%
  mutate(DistanceToCenter = distances)

# Step 4: Map Cluster number to archetype name
# Use the mapping from initial_centers
cluster_map <- initial_centers %>%
  mutate(Cluster = 1:n()) %>%
  select(Cluster, ARCHETYPE)

df_with_clusters_named <- df_with_clusters %>%
  left_join(cluster_map, by = "Cluster")

# Step 5: Filter out seed players
df_nonseeds <- df_with_clusters_named %>%
  anti_join(seed_keys, by = c("PLAYER_NAME", "SEASON"))

# Step 6: Get top 10 closest players per cluster
top10_by_cluster <- df_nonseeds %>%
  group_by(Cluster, ARCHETYPE) %>%
  slice_min(order_by = DistanceToCenter, n = 10) %>%
  arrange(Cluster, DistanceToCenter) %>%
  select(Cluster, ARCHETYPE, PLAYER_NAME, SEASON, DistanceToCenter)

# View the result
print(top10_by_cluster, n = 100)
```

```{r}
# Step 1: Join raw PTS back into clustered data
df_with_clusters_named_pts <- df_with_clusters_named %>%
  left_join(
    df_merged_clean %>% select(PLAYER_NAME, SEASON, PTS_raw = PTS),
    by = c("PLAYER_NAME", "SEASON")
  )

# Step 2: Use PTS_raw to filter and display top 10 scoring archetypes
top10_scoring_representatives <- df_with_clusters_named_pts %>%
  filter(PTS_raw > 5) %>%
  group_by(ARCHETYPE) %>%
  slice_min(order_by = DistanceToCenter, n = 20) %>%
  arrange(ARCHETYPE, DistanceToCenter) %>%
  select(ARCHETYPE, PLAYER_NAME, SEASON, PTS_raw, DistanceToCenter)

print(top10_scoring_representatives, n = 100)
```

```{r}
# Step 1: Join raw PTS back into clustered data
df_with_clusters_named_pts <- df_with_clusters_named %>%
  left_join(
    df_merged_clean %>%
      select(PLAYER_NAME, SEASON, PTS_raw = PTS),
    by = c("PLAYER_NAME", "SEASON")
  )

# Step 2: Filter for 2023–24 and get top 10 scorers per archetype
top10_scorers_per_cluster <- df_with_clusters_named_pts %>%
  filter(SEASON == "2023-24") %>%
  filter(!is.na(PTS_raw)) %>%
  group_by(ARCHETYPE) %>%
  slice_max(order_by = PTS_raw, n = 20, with_ties = FALSE) %>%
  arrange(ARCHETYPE, desc(PTS_raw)) %>%
  select(ARCHETYPE, PLAYER_NAME, SEASON, PTS_raw)

# View the result
print(top10_scorers_per_cluster, n = 100)
```


## Accuracy on left-out players

```{r}
# Step 1: Build test set with true labels
test_set <- tribble(
  ~PLAYER_NAME, ~SEASON, ~TRUE_ARCHETYPE,
  "Julius Randle", "2019-20", "Post Scorer",
  "Domantas Sabonis", "2022-23", "Post Scorer",
  "Joel Embiid", "2022-23", "Post Scorer",
  "DeMarcus Cousins", "2017-18", "Post Scorer",
  "Alperen Sengun", "2023-24", "Post Scorer",

  "Brook Lopez", "2019-20", "Stretch Big",
  "Myles Turner", "2023-24", "Stretch Big",
  "Kevin Love", "2017-18", "Stretch Big",
  "Bobby Portis", "2018-19", "Stretch Big",
  "John Collins", "2020-21", "Stretch Big",

  "Rudy Gobert", "2021-22", "Roll & Cut Big",
  "Clint Capela", "2017-18", "Roll & Cut Big",
  "Steven Adams", "2020-21", "Roll & Cut Big",
  "Ian Mahinmi", "2015-16", "Roll & Cut Big",
  "Dereck Lively II", "2023-24", "Roll & Cut Big",

  "James Harden", "2018-19", "Shot Creator",
  "LeBron James", "2017-18", "Shot Creator",
  "Zach LaVine", "2021-22", "Shot Creator",
  "Paolo Banchero", "2023-24", "Shot Creator",
  "Kawhi Leonard", "2018-19", "Shot Creator",

  "Kyle Lowry", "2018-19", "Primary Ball-Handler",
  "Tyrese Haliburton", "2020-21", "Primary Ball-Handler",
  "Jeremy Lin", "2016-17", "Primary Ball-Handler",
  "Mike Conley", "2021-22", "Primary Ball-Handler",
  "Reggie Jackson", "2017-18", "Primary Ball-Handler",

  "Aaron Holiday", "2021-22", "Slasher",
  "Markelle Fultz", "2019-20", "Slasher",
  "Dante Exum", "2017-18", "Slasher",
  "Michael Carter-Williams", "2015-16", "Slasher",
  "Dennis Smith Jr.", "2022-23", "Slasher",

  "Jaylen Brown", "2016-17", "Finisher",
  "Hamidou Diallo", "2019-20", "Finisher",
  "Jarred Vanderbilt", "2021-22", "Finisher",
  "Tony Allen", "2015-16", "Finisher",
  "Precious Achiuwa", "2022-23", "Finisher",

  "JJ Redick", "2017-18", "Movement Shooter",
  "Klay Thompson", "2021-22", "Movement Shooter",
  "Kyle Korver", "2016-17", "Movement Shooter",
  "Wesley Matthews", "2017-18", "Movement Shooter",
  "Joe Harris", "2017-18", "Movement Shooter",

  "P.J. Tucker", "2017-18", "Stationary Shooter",
  "Nicolas Batum", "2021-22", "Stationary Shooter",
  "Dorian Finney-Smith", "2022-23", "Stationary Shooter",
  "Royce O'Neale", "2020-21", "Stationary Shooter",
  "Dean Wade", "2023-24", "Stationary Shooter",
  
  "Andrew Wiggins", "2023-24", "2-Level Playmaker",
  "Bojan Bogdanovic", "2022-23", "2-Level Playmaker",
  "RJ Barrett", "2023-24", "2-Level Playmaker",
  "Jaylen Brown", "2017-18", "2-Level Playmaker",
  "Rui Hachimura", "2019-20", "2-Level Playmaker"
)

# Step 2: Merge with clustering results
test_predictions <- test_set %>%
  left_join(df_with_clusters_named %>%
              select(PLAYER_NAME, SEASON, Cluster, PREDICTED_ARCHETYPE = ARCHETYPE),
            by = c("PLAYER_NAME", "SEASON"))

# Step 3: Compute overall accuracy
overall_accuracy <- mean(test_predictions$TRUE_ARCHETYPE == test_predictions$PREDICTED_ARCHETYPE)
cat("✅ Overall Accuracy:", round(overall_accuracy * 100, 1), "%\n\n")

# Step 4: Compute accuracy per archetype
accuracy_by_archetype <- test_predictions %>%
  group_by(TRUE_ARCHETYPE) %>%
  summarise(
    correct = sum(TRUE_ARCHETYPE == PREDICTED_ARCHETYPE),
    total = n(),
    accuracy = round(100 * correct / total, 1),
    .groups = "drop"
  )

print(accuracy_by_archetype)

# Step 5: Print all test predictions (for inspection)
test_predictions %>%
  select(PLAYER_NAME, SEASON, TRUE_ARCHETYPE, PREDICTED_ARCHETYPE)
```

## Summary of a Player
```{r}
find_archetype_and_similar_players <- function(player_name, season, df_with_clusters_named, k = 10) {
  # Step 1: Get the target player row
  target_row <- df_with_clusters_named %>%
    filter(PLAYER_NAME == player_name, SEASON == season)

  if (nrow(target_row) != 1) {
    stop("Player-season not found or not unique.")
  }

  target_cluster <- target_row$Cluster
  target_archetype <- target_row$ARCHETYPE

  # Step 2: Get all players in same archetype (same cluster)
  same_archetype_df <- df_with_clusters_named %>%
    filter(Cluster == target_cluster)

  # Step 3: Compute similarity (Euclidean distance) to all others
  target_vec <- target_row %>%
    select(where(is.numeric)) %>%
    select(-Cluster, -DistanceToCenter)

  feature_matrix <- same_archetype_df %>%
    select(where(is.numeric)) %>%
    select(-Cluster, -DistanceToCenter)

  # Step 4: Use FNN to get neighbors
  knn_result <- get.knnx(
    data = feature_matrix,
    query = as.matrix(target_vec),
    k = k + 1  # include self
  )

  neighbor_indices <- knn_result$nn.index[1, ]
  neighbors <- same_archetype_df[neighbor_indices, ] %>%
    filter(!(PLAYER_NAME == player_name & SEASON == season)) %>%
    slice_head(n = k) %>%
    select(PLAYER_NAME, SEASON, ARCHETYPE, DistanceToCenter)

  cat(glue::glue("Player {player_name} ({season}) is a '{target_archetype}' archetype.\n\n"))
  return(neighbors)
}
```
```{r}
library(dplyr)
# See Cluster Overlap
get_cluster_similarity <- function(player_name, season, df_with_scaled, initial_centers, archetype_names) {
  # 1. Get player's standardized stats
  player_row <- df_with_scaled_mutated %>%
    filter(PLAYER_NAME == player_name, SEASON == season) %>%
    select(where(is.numeric))
  
  if (nrow(player_row) != 1) stop("Player-season not found or not unique.")
  
  # 2. Compute Euclidean distance from each cluster centroid
  distances <- apply(initial_centers %>% select(where(is.numeric)), 1, function(center) {
    sqrt(sum((player_row - center)^2))
  })
  
  # 3. Convert distances to similarity (inverse distance)
  max_dist <- max(distances)
  similarity <- (max_dist - distances) / max_dist
  
  # 4. Normalize to percentage
  similarity_pct <- similarity / sum(similarity) * 100
  
  # 5. Return as a tibble
  tibble(
    ARCHETYPE = archetype_names,
    SimilarityPercent = round(similarity_pct, 1)
  ) %>% arrange(desc(SimilarityPercent))
}
```
```{r}
get_player_archetype_history <- function(player_name, df_clustered) {
  df_clustered %>%
    filter(PLAYER_NAME == player_name) %>%
    arrange(SEASON) %>%
    select(PLAYER_NAME, SEASON, ARCHETYPE)
}
```
 
Enter Players Name:
```{r}
name = "Luka Doncic"
year = "2022-23"
```

```{r}
find_archetype_and_similar_players(name, year, df_with_clusters_named, k = 25)
```
```{r}
get_player_archetype_history(name, df_with_clusters_named)
```
```{r}
archetype_names <- initial_centers$ARCHETYPE
get_cluster_similarity(name, year, df_with_scaled_mutated, initial_centers, archetype_names)
```

## Plot Stats by Cluster
```{r}
# Step 1: Merge cluster labels with unscaled stats
df_clustered_merged_1 <- df_with_clusters_named %>%
  left_join(df_merged_clean %>%
              select(PLAYER_NAME, SEASON,
                     "PTS", "USG%",
                     "3P%FGA", "CORNER%3PA", "DUNK%FGA",
                     "Isolation_FREQ", "Spotup_FREQ"),
            by = c("PLAYER_NAME", "SEASON"))

# Step 2: Cluster averages for selected variables
cluster_summary_1 <- df_clustered_merged_1 %>%
  group_by(ARCHETYPE) %>%
  summarise(
    across(c("PTS.x", "USG%.x", "3P%FGA.x",
             "CORNER%3PA.x", "DUNK%FGA.x", "Isolation_FREQ.x", "Spotup_FREQ.x"),
           mean, na.rm = TRUE),
    .groups = "drop"
  )

# Step 3: Reshape and rename stats for plotting
cluster_long_2 <- cluster_summary_1 %>%
  pivot_longer(-ARCHETYPE, names_to = "Stat", values_to = "Mean") %>%
    mutate(
      Stat = recode(Stat,
        "PTS.x" = "Points",
        "USG%.x" = "Usage Rate",
        "3P%FGA.x" = "Three-Pointer Frequency",
        "CORNER%3PA.x" = "Corner Three-Pointer Frequency",
        "DUNK%FGA.x" = "Dunk Frequency",
        "Isolation_FREQ.x" = "Isolation Frequency",
        "Spotup_FREQ.x" = "Spotup Frequency"
      )
    )
```
```{r}
library(dplyr)
library(ggplot2)
library(tidyr)

# Step 4: Plot (Zoomed Out + Gridlines)
ggplot(cluster_long_2, aes(x = Mean, y = Stat, color = ARCHETYPE)) +
  geom_point(size = 3) +
  scale_color_viridis_d(option = "H") + 
  labs(
    title = "How Each Cluster Compares",
    x = "Cluster Average",
    y = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.major.y = element_line(color = "grey80"),
    panel.grid.major.x = element_line(color = "grey90"),
    legend.position = "right",  # Move legend to the right
    plot.margin = margin(10, 20, 10, 20),
    axis.text.y = element_text(size = 8)
  )

```

## Minimum Spanning Tree (2023-24)

```{r}
library(dplyr)
library(igraph)
library(tibble)
library(RColorBrewer)

# Step 1: Filter top players (excluding problematic names)
top_players <- df_with_clusters_named_pts %>%
  filter(SEASON == "2023-24") %>%
  filter(!(PLAYER_NAME %in% c(
    "Jerami Grant", "Miles Bridges", "Chet Holmgren", "RJ Barrett",
    "Malaki Branham", "Lonnie Walker IV", "Franz Wagner", "Mikal Bridges",
    "Jalen Green", "Zach LaVine", "Devin Vassell", "Tyler Herro",
    "Dalano Banton", "Andrew Nembhard", "De'Andre Hunter"
  ))) %>%
  group_by(ARCHETYPE) %>%
  slice_max(order_by = PTS_raw, n = 25) %>%
  ungroup() %>%
  select(PLAYER_NAME, SEASON, ARCHETYPE, DistanceToCenter, PTS_raw)

# Step 2: Scaled feature matrix for selected players
player_features <- df_with_scaled_mutated %>%
  semi_join(top_players, by = c("PLAYER_NAME", "SEASON")) %>%
  select(PLAYER_NAME, where(is.numeric)) %>%
  column_to_rownames("PLAYER_NAME")

# Step 3: Pairwise distances and edges
dist_matrix <- dist(player_features)
edges <- as.data.frame(as.table(as.matrix(dist_matrix))) %>%
  filter(Var1 != Var2) %>%
  rename(from = Var1, to = Var2, weight = Freq) %>%
  rowwise() %>%
  mutate(pair = paste(sort(c(from, to)), collapse = "_")) %>%
  distinct(pair, .keep_all = TRUE) %>%
  select(from, to, weight)

# Step 4: Node metadata
nodes <- top_players %>%
  filter(PLAYER_NAME %in% unique(c(edges$from, edges$to))) %>%
  distinct(PLAYER_NAME, ARCHETYPE) %>%
  rename(name = PLAYER_NAME, group = ARCHETYPE)

# Step 5: Build graph and MST
g <- graph_from_data_frame(edges, vertices = nodes, directed = FALSE)
mst <- mst(g, weights = E(g)$weight)

# Step 6: Build layout
set.seed(42)
layout <- layout_with_kk(mst)  # or layout_with_kk(mst)
layout <- layout * 16

# Step 7: Build label vector: only top scorer per archetype
top_labeled <- top_players %>%
  group_by(ARCHETYPE) %>%
  arrange(desc(PTS_raw)) %>%
  filter(row_number() %in% c(1, 3)) %>%
  pull(PLAYER_NAME)

# Highlight vertex frame for labeled players
V(mst)$frame.color <- ifelse(
  V(mst)$name %in% top_labeled,
  "firebrick",       # highlight labeled players with a black border
  NA             # default (invisible) border
)
V(mst)$label <- ifelse(V(mst)$name %in% top_labeled, V(mst)$name, NA)

# Step 8: Color palette for clusters
n_clusters <- length(unique(V(mst)$group))
cluster_colors <- colorRampPalette(brewer.pal(8, "Dark2"))(n_clusters)
vertex_colors <- cluster_colors[as.numeric(as.factor(V(mst)$group))]

par(mar = c(0, 0, 1, 0), family = "Tahoma")
# Step 9: Plot MST
plot(
  mst,
  layout = layout,
  main = "2023–24 NBA Archetype MST",
  vertex.label = V(mst)$label,
  vertex.label.cex = 0.6,   # Adjust font size of labels
  vertex.label.color = "black",
  vertex.label.dist = 1.1,
  #vertex.label.degree = runif(vcount(mst), 0, pi),
  # Adjust distance of labels from vertices (lower for closer to center)
  #vertex.label.family = "sans",  # Font family for labels
  vertex.size = 5,         # Increase vertex size for better visibility
  edge.width = 4,           # Increase edge width for better visibility
  edge.color = adjustcolor("gray60", alpha.f = 0.7),
  vertex.color = vertex_colors
)

# Add legend for clusters
legend("bottomleft",
       legend = levels(as.factor(V(mst)$group)),
       fill = cluster_colors,
       title = "Archetype",
       cex = 0.8,
       bty = "n")

```

## Past NBA Champions:
```{r}
library(dplyr)
library(ggplot2)

# Define the lineups for each team
championship_lineups <- tribble(
  ~TEAM, ~PLAYER_NAME, ~SEASON,
  
  # 2024 Celtics
  "24 Celtics", "Al Horford", "2023-24",
  "24 Celtics", "Jrue Holiday", "2023-24",
  "24 Celtics", "Jayson Tatum", "2023-24",
  "24 Celtics", "Jaylen Brown", "2023-24",
  "24 Celtics", "Derrick White", "2023-24",
  
  # 2023 Nuggets
  "23 Nuggets", "Nikola Jokic", "2022-23",
  "23 Nuggets", "Jamal Murray", "2022-23",
  "23 Nuggets", "Michael Porter Jr.", "2022-23",
  "23 Nuggets", "Aaron Gordon", "2022-23",
  "23 Nuggets", "Kentavious Caldwell-Pope", "2022-23",
  
  # 2022 Warriors
  "22 Warriors", "Stephen Curry", "2021-22",
  "22 Warriors", "Klay Thompson", "2021-22",
  "22 Warriors", "Draymond Green", "2021-22",
  "22 Warriors", "Andrew Wiggins", "2021-22",
  "22 Warriors", "Kevon Looney", "2021-22",
  
  # 2021 Bucks
  "21 Bucks", "P.J. Tucker", "2020-21",
  "21 Bucks", "Brook Lopez", "2020-21",
  "21 Bucks", "Jrue Holiday", "2020-21",
  "21 Bucks", "Khris Middleton", "2020-21",
  "21 Bucks", "Giannis Antetokounmpo", "2020-21",
  
  # 2020 Lakers
  "20 Lakers", "LeBron James", "2019-20",
  "20 Lakers", "JaVale McGee", "2019-20",
  "20 Lakers", "Danny Green", "2019-20",
  "20 Lakers", "Anthony Davis", "2019-20",
  "20 Lakers", "Kentavious Caldwell-Pope", "2019-20",
  
  # 2019 Raptors
  "19 Raptors", "Kyle Lowry", "2018-19",
  "19 Raptors", "Marc Gasol", "2018-19",
  "19 Raptors", "Danny Green", "2018-19",
  "19 Raptors", "Kawhi Leonard", "2018-19",
  "19 Raptors", "Pascal Siakam", "2018-19",
  
  # 2018 Warriors
  "18 Warriors", "Andre Iguodala", "2017-18",
  "18 Warriors", "Kevin Durant", "2017-18",
  "18 Warriors", "Stephen Curry", "2017-18",
  "18 Warriors", "Klay Thompson", "2017-18",
  "18 Warriors", "Draymond Green", "2017-18",
  
  # 2017 Cavs
  "17 Cavs", "LeBron James", "2016-17",
  "17 Cavs", "JR Smith", "2016-17",
  "17 Cavs", "Kevin Love", "2016-17",
  "17 Cavs", "Kyrie Irving", "2016-17",
  "17 Cavs", "Tristan Thompson", "2016-17",
  
  # 2016 Warriors
  "16 Warriors", "Andre Iguodala", "2015-16",
  "16 Warriors", "Stephen Curry", "2015-16",
  "16 Warriors", "Klay Thompson", "2015-16",
  "16 Warriors", "Harrison Barnes", "2015-16",
  "16 Warriors", "Draymond Green", "2015-16"
)

```

```{r}
library(dplyr)
library(ggplot2)
library(stringr)

# Step 1: Add last names to each player
lineup_archetypes_labeled <- championship_lineups %>%
  left_join(df_with_clusters_named, by = c("PLAYER_NAME", "SEASON")) %>%
  mutate(
    LAST_NAME = word(PLAYER_NAME, -1),  # Extract last name
      LAST_NAME = case_when(
        PLAYER_NAME == "Kentavious Caldwell-Pope" ~ "C-Pope",
        PLAYER_NAME == "Michael Porter Jr." ~ "Porter Jr.",
        PLAYER_NAME == "Giannis Antetokounmpo" ~ "Giannis",
        TRUE ~ LAST_NAME
      )
  )

# Step 2: Plot
ggplot(lineup_archetypes_labeled, aes(x = TEAM, fill = ARCHETYPE)) +
  geom_bar(position = "stack", stat = "identity", aes(y = 1)) +  # One bar per player
  geom_text(aes(y = 1, label = LAST_NAME), 
            position = position_stack(vjust = 0.5), 
            size = 3, color = "white") +
  coord_flip() +
  labs(
    title = "Breakdown of Most-Used Lineups (Past Champions)",
    x = "Team",
    y = "Number of Players",
    fill = "Archetype"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

```{r}
library(dplyr)
library(ggplot2)
library(stringr)
library(tibble)

# Define All-NBA First Teams data
all_nba_first_teams <- tribble(
  ~TEAM,                  ~PLAYER_NAME,               ~SEASON,
  "23-24 ",     "Giannis Antetokounmpo",    "2023-24",
  "23-24 ",     "Luka Doncic",              "2023-24",
  "23-24 ",     "Shai Gilgeous-Alexander",  "2023-24",
  "23-24 ",     "Nikola Jokic",             "2023-24",
  "23-24 ",     "Jayson Tatum",             "2023-24",

  "22-23 ",     "Giannis Antetokounmpo",    "2022-23",
  "22-23 ",     "Jayson Tatum",             "2022-23",
  "22-23 ",     "Joel Embiid",              "2022-23",
  "22-23 ",     "Shai Gilgeous-Alexander",  "2022-23",
  "22-23 ",     "Luka Doncic",              "2022-23",

  "21-22 ",     "Giannis Antetokounmpo",    "2021-22",
  "21-22 ",     "Jayson Tatum",             "2021-22",
  "21-22 ",     "Nikola Jokic",             "2021-22",
  "21-22 ",     "Devin Booker",             "2021-22",
  "21-22 ",     "Luka Doncic",              "2021-22",

  "20-21 ",     "Giannis Antetokounmpo",    "2020-21",
  "20-21 ",     "Kawhi Leonard",            "2020-21",
  "20-21 ",     "Nikola Jokic",             "2020-21",
  "20-21 ",     "Stephen Curry",            "2020-21",
  "20-21 ",     "Luka Doncic",              "2020-21",

  "19-20 ",     "Giannis Antetokounmpo",    "2019-20",
  "19-20 ",     "LeBron James",             "2019-20",
  "19-20 ",     "Anthony Davis",            "2019-20",
  "19-20 ",     "James Harden",             "2019-20",
  "19-20 ",     "Luka Doncic",              "2019-20",

  "18-19 ",     "Giannis Antetokounmpo",    "2018-19",
  "18-19 ",     "Paul George",              "2018-19",
  "18-19 ",     "Nikola Jokic",             "2018-19",
  "18-19 ",     "James Harden",             "2018-19",
  "18-19 ",     "Stephen Curry",            "2018-19",

  "17-18 ",     "Kevin Durant",             "2017-18",
  "17-18 ",     "LeBron James",             "2017-18",
  "17-18 ",     "Anthony Davis",            "2017-18",
  "17-18 ",     "James Harden",             "2017-18",
  "17-18 ",     "Damian Lillard",           "2017-18",

  "16-17 ",     "Kawhi Leonard",            "2016-17",
  "16-17 ",     "LeBron James",             "2016-17",
  "16-17 ",     "Anthony Davis",            "2016-17",
  "16-17 ",     "James Harden",             "2016-17",
  "16-17 ",     "Russell Westbrook",        "2016-17",

  "15-16 ",     "Kawhi Leonard",            "2015-16",
  "15-16 ",     "LeBron James",             "2015-16",
  "15-16 ",     "DeAndre Jordan",           "2015-16",
  "15-16 ",     "Stephen Curry",            "2015-16",
  "15-16 ",     "Russell Westbrook",        "2015-16"
)

# Annotate with archetypes and last names
all_nba_labeled <- all_nba_first_teams %>%
  left_join(df_with_clusters_named, by = c("PLAYER_NAME","SEASON")) %>%
  mutate(
    LAST_NAME = word(PLAYER_NAME, -1),
    LAST_NAME = case_when(
      PLAYER_NAME == "Giannis Antetokounmpo" ~ "Giannis",
      PLAYER_NAME == "Shai Gilgeous-Alexander" ~ "SGA",
      TRUE ~ LAST_NAME
    ),
    ARCHETYPE = if_else(PLAYER_NAME == "Luka Doncic", "Shot Creator", ARCHETYPE)
  )

# Plot horizontal stacked bar chart with last name labels
ggplot(all_nba_labeled, aes(x = TEAM, fill = ARCHETYPE)) +
  geom_bar(stat = "identity", aes(y = 1), position = "stack", width = 0.7) +
  geom_text(aes(y = 1, label = LAST_NAME),
            position = position_stack(vjust = 0.5),
            size = 3, color = "white") +
  coord_flip() +
  labs(
    title = "",
    x = "Season All-NBA First Team",
    y = "Number of Players",
    fill = "Archetype"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")

```




# Initial K-Means Clustering (Just Play Types)
```{r}
library(dplyr)

# Select only numeric columns to cluster on
df_numeric <- df_wide %>%
  select(where(is.numeric)) %>%
  select(-PLAYER_ID)  # remove identifier columns if needed

# Save player-season info to merge back later
player_info <- df_wide %>%
  select(SEASON, PLAYER_NAME, PLAYER_ID)

# If NA, use mean of that column (Mess with this !!)
df_numeric_clean <- df_numeric %>%
  mutate(across(everything(), ~ ifelse(is.finite(.), ., mean(., na.rm = TRUE))))

# Normalize the numeric values
df_scaled <- scale(df_numeric_clean)
```


```{r}
set.seed(42)  # for reproducibility
k <- 8
kmeans_result <- kmeans(df_scaled, centers = k, nstart = 25)
```

```{r}
df_archetypes <- bind_cols(player_info, as.data.frame(df_scaled)) %>%
  mutate(Archetype = as.factor(kmeans_result$cluster))
```

```{r}
df_archetypes %>%
  count(Archetype) %>%
  arrange(desc(n))

```

```{r}
df_archetypes %>%
  select(SEASON, PLAYER_NAME, Archetype) %>%
  arrange(Archetype, SEASON, PLAYER_NAME)

```

```{r}
df_archetypes %>%
  group_by(Archetype) %>%
  summarise(Players = paste0(unique(PLAYER_NAME), collapse = ", "), .groups = "drop")

```

